---
title: "WNBA-technical-fouls"
output: pdf_document
---

```{r}
library(tidyverse)
library(brms)
library(tidybayes)
'%!in%' <- function(x,y)!('%in%'(x,y))
```

```{r}
data = read.csv("espn.com/espn-pbp-1997-2021-v2.csv")
game_info = read.csv('espn.com/game_info.csv')
```

```{r}
# join to get the year info and eventually other game info if needed
data = data %>% left_join(game_info)
data = data %>% mutate(year = as.factor(year))

# table of number of total games per year
data %>% 
  dplyr::select(GameID,year) %>% 
  unique() %>% group_by(year) %>% 
  dplyr::summarise(num_games=n())

data = data %>% filter(year!=2002 & year!=2022) %>% # 2002 not enough games
  filter(GameType != "preseason") %>% # remove preseason for this analysis
  filter(!(HomeTeam == "EAST"|HomeTeam=="WEST")) # remove all-star game

# get total number of games per year again to compute league avg. technical rate later
games_per_year = data %>% 
  dplyr::select(GameID,year) %>% 
  unique() %>% group_by(year) %>% 
  dplyr::summarise(num_games=n())
```




# Team name cleaning
```{r}
# print all unique teams
data$EventTeam %>% unique()

data = data %>% mutate(EventTeam = ifelse(EventTeam=="WSH","WAS",
                                          ifelse(EventTeam=="CT","CONN",
                                                 ifelse(EventTeam=="NY","NYL",
                                                        ifelse(EventTeam=="PHO","PHX",
                                                               ifelse(EventTeam=="LOS","LA",
                                                                      ifelse(EventTeam=="SA","SAS",EventTeam)))))))

data = data %>% mutate(AwayTeam = ifelse(AwayTeam=="WSH","WAS",
                                          ifelse(AwayTeam=="CT","CONN",
                                                 ifelse(AwayTeam=="NY","NYL",
                                                        ifelse(AwayTeam=="PHO","PHX",
                                                               ifelse(AwayTeam=="LOS","LA",
                                                                      ifelse(AwayTeam=="SA","SAS",AwayTeam)))))))

data = data %>% mutate(HomeTeam = ifelse(HomeTeam=="WSH","WAS",
                                          ifelse(HomeTeam=="CT","CONN",
                                                 ifelse(HomeTeam=="NY","NYL",
                                                        ifelse(HomeTeam=="PHO","PHX",
                                                               ifelse(HomeTeam=="LOS","LA",
                                                                      ifelse(HomeTeam=="SA","SAS",HomeTeam)))))))

# get number of games per team in order to normalize technical rates later
games_per_year_by_team = data %>% 
  dplyr::select(GameID,year,EventTeam) %>% 
  unique() %>% 
  group_by(year,EventTeam) %>% 
  dplyr::summarise(num_games=n())
games_per_year_by_team

# get list of games for poisson model for when there were no technicals in a game
all_games = data %>% dplyr::select(GameID,year,EventTeam) %>% unique()
```

```{r}
# did DT play? use the PBP to deduce whether DT played at all. This will be a variable in the model later
did_dt_play = data %>% mutate(DT_play = grepl("Taurasi",EventDescription)) %>% group_by(GameID) %>% dplyr::summarise(DT_played = max(DT_play))
```


```{r}
# now use regex to get the PBP rows that have technical fouls
data = data %>% mutate(tech = grepl("[Tt]ech",EventDescription))

# is there anything other than tech fouls or free throws?
data %>% filter(tech) %>% 
  filter(!grepl("[Tt]ech.*cal [Ff]ree [Tt]hrow",EventDescription))  %>% # remove technical free throws
  filter(!grepl("[Ff]oul",EventDescription))
```

# Clean double techs by adding a second row
```{r}
# filter data to all techs, removing the free throws
techs = data %>% filter(tech) %>% 
  filter(!grepl("[Tt]ech.*cal [Ff]ree [Tt]hrow",EventDescription))

# new column to track who the technical foul was against
techs$TechPlayer = NA

# print all of the double technicals so we can see the patterns in the descriptions
techs %>% filter(grepl("[Dd]ouble",EventDescription))

# assign the name of the first player
techs = techs %>% mutate(TechPlayer = ifelse(grepl("[Dd]ouble",EventDescription), # assign for pattern 1
                                     gsub(" and.*","",gsub("Double [Tt]echnical [Ff]oul: ","",EventDescription)),
                                     NA)) %>% 
  mutate(TechPlayer = ifelse(grepl("gains possession",EventDescription), # assign for pattern 2
                                     gsub(" Double Technical.*","",EventDescription),
                                     TechPlayer))


# assign the name of the second player
doubles = techs %>% filter(grepl("[Dd]ouble",EventDescription)) %>% 
  mutate(EventTeam = ifelse(EventTeam==HomeTeam,AwayTeam,HomeTeam)) %>% # swap the team who the foul is against
  mutate(HomeOrAway = ifelse(HomeOrAway=="home","away","home"))  %>%  # swap home and away
  mutate(TechPlayer = ifelse(grepl("[Dd]ouble",EventDescription), # assign for pattern 1
                             gsub(".*and ","",EventDescription),
                             NA)) %>%
  mutate(TechPlayer = ifelse(grepl("gains possession",EventDescription), # assign for pattern 2
                                     gsub("gains possession\\.","",
                                          gsub(".*Double Technical Foul\\. ","",EventDescription)),
                                     TechPlayer))

# now bind on those additional rows
techs = rbind(techs,doubles)


# get the player who committed the tech for all techs that weren't double techs
techs = techs %>% 
  mutate(TechPlayer = ifelse(grepl("[Dd]ouble",EventDescription),
                             TechPlayer,
                             gsub(" ?(flopping|indirect)? ?[Tt]echnical [Ff]oul.*| ?[Dd]efensive.*| [Nn]on.*| [Tt]aunting [Tt]echnical.*","",EventDescription))) %>% 
  mutate(TechPlayer = ifelse(TechPlayer %in% c("delay techfoul","Delay of Game Technical","- Bench","non unsport techfoul","Too Many Players Technical","excess timeout techfoul"),"",TechPlayer)) # some weird PBP entries
```

```{r}
# label the defensive 3-second techs
techs = techs %>% mutate(TechType = ifelse(grepl("[Dd]efensive",EventDescription),"Defensive 3-Second","Other"))
```


# All-time technical fouls by franchise
```{r}
techs %>% ggplot(aes(x=EventTeam)) + geom_bar()

#same plot but exclude the old franchises
techs %>% filter(EventTeam %!in% c("SA","SAS","WEST","EAST","TUL","HOU","CHA","SAC")) %>%
  ggplot(aes(x=EventTeam)) + geom_bar()
```


```{r}
# plot technicals by year: the big jump is due to the defensive 3-second techs
techs %>% ggplot(aes(x=year,fill=TechType)) + geom_bar()

techs %>% group_by(year) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year) %>% 
  mutate(techs_per_game = num_techs/num_games) %>%
  ggplot(aes(x=year,y=techs_per_game)) + 
  geom_col()
```

# Not Defensive 3-Second 
```{r}
# excluding defensive 3-second techs
non_def_techs = techs %>% filter(TechType != "Defensive 3-Second")

# number of techs per year excluding defensive 3-second techs
non_def_techs %>% ggplot(aes(x=year)) + geom_bar()

# same plot normalized for number of games
non_def_techs %>% group_by(year) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year) %>% 
  mutate(techs_per_game = num_techs/num_games) %>%
  ggplot(aes(x=year,y=techs_per_game)) + 
  geom_col()
```
# Defensive 3-Second Technicals
```{r}
def_techs = techs %>% filter(TechType=="Defensive 3-Second") 

def_techs %>% ggplot(aes(x=year,fill=GameType)) + geom_bar() + facet_wrap(~EventTeam)

def_techs %>% ggplot(aes(x=year,fill=TechType)) + geom_bar()

def_techs %>% group_by(year) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year) %>% #mutate(year = as.numeric(as.character(year))) %>%
  mutate(techs_per_game = num_techs/num_games) %>%
  ggplot(aes(x=year,y=techs_per_game)) + 
  geom_col()
```

```{r}
non_def_techs %>% group_by(year) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year) %>% mutate(year = as.numeric(as.character(year))) %>%
  mutate(techs_per_game = num_techs/num_games) %>%
  ggplot(aes(x=year,y=techs_per_game)) + 
  geom_line()

league_avg = non_def_techs %>% group_by(year) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year) %>% mutate(year = as.numeric(as.character(year))) %>%
  mutate(techs_per_game = num_techs/num_games)

non_def_techs %>% ggplot(aes(x=EventTeam)) + geom_bar()
non_def_techs %>% filter(EventTeam %!in% c("SA","SAS","WEST","EAST","TUL","HOU","CHA","SAC")) %>% 
  ggplot(aes(x=EventTeam)) + geom_bar()

non_def_techs %>% ggplot(aes(x=year,fill=GameType)) + geom_bar() + facet_wrap(~EventTeam,) + scale_x_discrete(breaks=c(2003,2012,2021)) + ggtitle("Number of technical fouls per team by year\n2003-2021 (Excluding def 3-sec techs)")
```

```{r}
# plot tech rate over time for each team all on the same plot
non_def_techs %>% group_by(year,EventTeam) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year_by_team) %>% mutate(year = as.numeric(as.character(year))) %>%
  mutate(techs_per_game = num_techs/num_games) %>%
  ggplot(aes(x=year,y=techs_per_game,col=EventTeam)) + 
  geom_line()

# now in boxplot form so you can see the league trends better
non_def_techs %>% group_by(year,EventTeam) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year_by_team) %>% 
  mutate(techs_per_game = num_techs/num_games) %>%
  ggplot(aes(x=year,y=techs_per_game)) + 
  geom_boxplot()
```


```{r}
league_avg_to_subtract = league_avg %>% 
  dplyr::select(year,techs_per_game) %>% 
  dplyr::rename(league_avg=techs_per_game) %>% mutate(league_avg=league_avg/2)

# plot difference from league average all on one plot
non_def_techs %>% group_by(year,EventTeam) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year_by_team) %>% 
  mutate(techs_per_game = num_techs/num_games) %>% ungroup() %>%
  mutate(year = as.numeric(as.character(year))) %>%
  left_join(league_avg_to_subtract) %>%
  mutate(diff_from_league_avg = techs_per_game-league_avg) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  ggplot(aes(x=year,y=diff_from_league_avg,col=EventTeam)) + 
  geom_line()

# plot difference for each team with the delta highlighted
non_def_techs %>% group_by(year,EventTeam) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year_by_team) %>% 
  mutate(techs_per_game = num_techs/num_games) %>% ungroup() %>%
  mutate(year = as.numeric(as.character(year))) %>%
  left_join(league_avg_to_subtract) %>%
  mutate(diff_from_league_avg = techs_per_game-league_avg) %>%
  ggplot(aes(x=year,y=diff_from_league_avg)) + 
  geom_col(aes(fill = diff_from_league_avg <0)) + 
  facet_wrap(~EventTeam) + theme(legend.position = "none") +
  ggtitle("Rate of technical fouls per game by team relative to league average\n2003-2021 (Excluding def 3-sec techs)")
```





# Diana Taurasi Technical Fouls

```{r}
# who are the league leaders in all-time technicals
non_def_techs %>% group_by(TechPlayer) %>% 
  dplyr::summarize(num_techs=n()) %>% 
  filter(TechPlayer!="") %>% 
  arrange(desc(num_techs)) %>% dplyr::rename("Career Technical Fouls (2003-2021)"=num_techs,
                                             Player = TechPlayer)
```

```{r}
# count of techs per year
non_def_techs %>% filter(TechPlayer == "Diana Taurasi") %>% 
  #mutate(year = as.numeric(as.character(year))) %>%
  ggplot(aes(x=year)) + 
  geom_bar(fill="orange") + ggtitle("Diana Taurasi Technical Fouls by Year")

# dt share of mercury technical fouls
non_def_techs %>% filter(EventTeam == "PHX") %>% filter(year!="2003") %>%
  mutate(EventTeam = ifelse(TechPlayer=="Diana Taurasi","Diana Taurasi","a")) %>%
  #mutate(year = as.numeric(as.character(year))) %>%
  ggplot(aes(x=year)) + 
  geom_bar(aes(fill=EventTeam)) +
  #geom_bar(data=subset(non_def_techs,TechPlayer == "Diana Taurasi"),fill="orange")  + 
  ggtitle("Diana Taurasi share of Mercury technical fouls by year") + theme_bw() + 
  scale_fill_manual(name = "Technical Against",values=c("purple","orange"),labels=c("Other Mercury Players","Diana Taurasi")) + 
  theme(legend.position = "bottom")
```
https://www.basketball-reference.com/wnba/players/t/tauradi01w.html
```{r}
# use external data for the number of minutes played
dt = read.csv("taurasi-minutes-played.csv") %>% dplyr::select(Year,G,MP) 
names(dt) = c("year","num_games","minutes_per_game")
dt = dt %>% mutate(year = as.factor(year))

# filter dataset to only dt fouls for some plotting
dt_data = non_def_techs %>% filter(TechPlayer == "Diana Taurasi") %>% 
  group_by(year) %>% 
  summarize(num_techs = n()) %>% 
  left_join(dt) %>% mutate(year = as.numeric(as.character(year))) %>%
  mutate(total_mins_played = num_games*minutes_per_game)
```



```{r}
# techs per games played by year 
dt_data %>% 
  ggplot(aes(x=year,y=num_techs/num_games)) + 
  geom_col(fill="orange") + ggtitle("Diana Taurasi technical fouls per games played")

# techs per total minutes played by year 
dt_data %>%
  ggplot(aes(x=year,y=num_techs/(num_games*minutes_per_game))) + 
  geom_col(fill="orange") + ggtitle("Diana Taurasi Technicals per Minutes Played")
```


TO DO: delete the years that Taurasi didn't play from these plots
```{r}
non_def_techs %>% mutate(EventTeam = ifelse(TechPlayer=="Diana Taurasi","Taurasi",EventTeam)) %>% ggplot(aes(x=year,fill=GameType)) + geom_bar() + facet_wrap(~EventTeam)

# get tech rate and diff from league rate for DT
dt_temp = dt_data %>% dplyr::select(-minutes_per_game,-total_mins_played) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  mutate(EventTeam = "Taurasi") %>% mutate(techs_per_game = num_techs/num_games) %>% ungroup() %>%
  left_join(league_avg_to_subtract) %>%
  mutate(diff_from_league_avg = techs_per_game-league_avg)

# get tech rate and diff from league rate for mercury minus DT
temp = non_def_techs %>% filter(TechPlayer!="Diana Taurasi") %>%
  group_by(year,EventTeam) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year_by_team) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  mutate(techs_per_game = num_techs/num_games) %>% ungroup() %>%
  left_join(league_avg_to_subtract) %>%
  mutate(diff_from_league_avg = techs_per_game-league_avg)

# get tech rate and diff from league rate for everyone else
temp2 = non_def_techs %>%
  group_by(year,EventTeam) %>% summarize(num_techs = n()) %>% 
  left_join(games_per_year_by_team) %>%
  mutate(year = as.numeric(as.character(year))) %>%
  mutate(techs_per_game = num_techs/num_games) %>% ungroup() %>%
  left_join(league_avg_to_subtract) %>%
  mutate(diff_from_league_avg = techs_per_game-league_avg)

# tech foul rate per game for DT vs Mercury
temp %>%
  ggplot(aes(x=year,y=techs_per_game)) + 
  geom_line(aes(group=EventTeam),subset(temp,EventTeam!="PHX"),alpha=.2) + 
  geom_line(aes(x=year,y=techs_per_game),subset(temp,EventTeam=="PHX"),col="purple",linetype="dashed") + 
  geom_line(aes(x=year,y=techs_per_game),subset(temp2,EventTeam=="PHX"),col="purple") + 
  geom_line(aes(x=year,y=techs_per_game/2),data=league_avg) + 
  geom_line(aes(x=year,y=techs_per_game),data=dt_temp,col="orange")

# difference from the league avg for DT vs Mercury
temp %>%
  ggplot(aes(x=year,y=diff_from_league_avg)) + 
  geom_line(aes(group=EventTeam),subset(temp,EventTeam!="PHX"),alpha=.2) + 
  geom_line(aes(x=year,y=diff_from_league_avg),subset(temp,EventTeam=="PHX"),col="purple",linetype="dashed",cex=1.5) + 
  geom_line(aes(x=year,y=diff_from_league_avg),subset(temp2,EventTeam=="PHX"),col="purple",cex=1.5) + 
  geom_line(aes(x=year,y=diff_from_league_avg),data=dt_temp,col="orange",cex=1.5) + 
  geom_abline(slope=0)
```

# DT Model
```{r}
dt_model = brm(num_techs~num_games,family=poisson(),data=subset(dt_data,year>=2010),chains=4,cores=4)
data.frame(num_games=30) %>% add_predicted_draws(dt_model) %>% 
  ungroup() %>%
  ggplot(aes(x=.prediction)) + 
  geom_histogram(binwidth = 1)
```

# Mercury Model
```{r}
team_data = techs %>%
  group_by(GameID,EventTeam,GameType) %>%
  dplyr::summarise(num_techs = n()) %>% 
  left_join(all_games,.) %>% 
  left_join(did_dt_play) %>% 
  mutate(DT_played = ifelse(is.na(DT_played),0,DT_played)) %>%
  mutate(num_techs = ifelse(is.na(num_techs),0,num_techs)) %>%
  mutate(year = as.numeric(as.character(year))) %>% 
  filter(EventTeam %!in% c("SA","SAS","WEST","EAST","TUL","HOU","CHA","SAC"))

merc_model = brm(num_techs~DT_played,family=poisson(),
                 data=subset(team_data,year>=2010 & EventTeam=="PHX"),chains=4,cores=4)
```

# Predictions
TO DO: work in uncertainty on the number of games
```{r}
data.frame(DT_played=c(rep(1,26),rep(0,10))) %>% add_predicted_draws(merc_model) %>% 
  ungroup() %>%
  group_by(.draw) %>%
  dplyr::summarise(season_total = sum(.prediction)) %>%
  ggplot(aes(x=season_total)) + 
  geom_histogram(binwidth = 1)

data.frame(DT_played=c(rep(1,26),rep(0,10))) %>% add_predicted_draws(merc_model) %>% 
  ungroup() %>%
  group_by(.draw) %>%
  dplyr::summarise(season_total = sum(.prediction)) %>%
  dplyr::summarise(mean(season_total))

data.frame(num_games=29) %>% add_predicted_draws(dt_model) %>% 
  ungroup() %>% dplyr::summarise(mean(.prediction))
```

```{r}
all_teams_model = brm(num_techs~1+(1|year)+(1|EventTeam),family=poisson(),
                 data=subset(team_data,year>=2010),chains=4,cores=4)

all_teams_model_DT = brm(num_techs~1+DT_played+(1|year)+(1+DT_played|EventTeam),family=poisson(),
                 data=subset(team_data,year>=2010),chains=4,cores=4)
```

```{r}
coef(all_teams_model_DT)
```

Who is getting technical fouls on Chicago when playing against Dt?
```{r}
techs %>%
  left_join(all_games,.) %>% 
  left_join(did_dt_play) %>% 
  mutate(DT_played = ifelse(is.na(DT_played),0,DT_played)) %>% 
  filter(DT_played==1 & EventTeam=="CHI" & tech) %>%
  group_by(TechPlayer) %>%
  dplyr::summarise(num_techs = n()) %>% arrange(desc(num_techs))
```

